---
title: "Analyzing Divvy Bike Usage Patterns in Chicago"
author: "Mohammad Hamza Piracha, Poshan Pandey, Usman Matheen"
date: "2024-07-04"
output:
  html_document: default
  pdf_document: default
---

# Data Preparation, Cleaning, and Transformation

## Data Preparation

### Installing Libraries

```{r}
library(isotree)
library(magrittr)
library(dplyr)
library(tidyr)
library(lubridate)
library(tidymodels)
library(ggplot2)
library(dbscan)
```

### Setting WD to dataset file and listing all dataset

```{r}
csv_files <- list.files(pattern = "*.csv")
```

### Reading and sorting dataframes

```{r}
data_list <- list()
for (file in csv_files) {
  file_path <- file.path("D:/Desktop/CSP-571-Proj/dataset", file)
  data <- read.csv(file_path)
  data_list[[file]] <- data
}
```

### Merging data for all months into 1

```{r}
merged_data <- do.call(rbind, data_list)
```

### Saving the merged data as merged_data

```{r}
write.csv(merged_data, "D:/Desktop/CSP-571-Proj/dataset/merged_data.csv", row.names = FALSE)
```

## Data cleaning and Transformation

### Data Cleaning

#### Removing Unnecessary Columns

```{r}
necessary <- c("ride_id","rideable_type", "started_at", "ended_at", "start_station_name", "end_station_name", "member_casual")
annual <- merged_data %>% select(all_of(necessary))
```

#### Removing Missing Values

```{r}
cleaned_annual <- annual %>% drop_na()
dim(cleaned_annual)
```

#### Convert 'started_at' and 'ended_at' to datetime

```{r}
cleaned_annual <- cleaned_annual %>%
  mutate(
    started_at = ymd_hms(started_at),
    ended_at = ymd_hms(ended_at)
  )
```

#### Create new features: trip duration, day of the week, hour of the day

```{r}
cleaned_annual <- cleaned_annual %>%
  mutate(
    trip_duration = as.numeric(difftime(ended_at, started_at, units = "mins")),
    day_of_week = wday(started_at, label = TRUE),
    start_hour = hour(started_at)
  )
```

#### Remove negative or zero trip durations

```{r}
cleaned_annual <- cleaned_annual %>%
  filter(trip_duration > 0)
```

#### Handle outliers in trip duration (e.g., removing trips longer than 24 hours)

```{r}
cleaned_annual <- cleaned_annual %>%
  filter(trip_duration <= 1440)
```

#### Check for consistency: 'ended_at' should be after 'started_at'

```{r}
cleaned_annual <- cleaned_annual %>%
  filter(ended_at > started_at)
dim(cleaned_annual)
```

#### Verifying cleaned data and saving it for further analysis

```{r}
dim(cleaned_annual)
summary(cleaned_annual)
write.csv(cleaned_annual, "D:/Desktop/CSP-571-Proj/dataset/cleaned_annual.csv", row.names = FALSE)
```

### Outlier Detection

#### Z-score Method for outlier detection

The Z-score method standardizes the dataset and identifies outliers based on a threshold, usually 3 standard deviations from the mean.

```{r}
# Calculate Z-scores
cleaned_annual <- cleaned_annual %>%
  mutate(z_score = (trip_duration - mean(trip_duration)) / sd(trip_duration))

# Filter out rows with Z-scores greater than 3 or less than -3
cleaned_annual <- cleaned_annual %>%
  filter(abs(z_score) <= 3) %>%
  select(-z_score)
```

#### Isolation Forest Method for outlier detection

Isolation Forest is a machine learning algorithm that isolates observations by randomly selecting a feature and then randomly selecting a split value between the maximum and minimum values of the selected feature.

```{r}
#### Isolation Forest Method for outlier detection
data_matrix <- as.matrix(cleaned_annual)

# Fit Isolation Forest model
iso_forest <- isolation.forest(data_matrix, ntrees = 100)

# Predict outliers
outlier_scores <- predict(iso_forest, data_matrix)

# Set a threshold for outlier scores
threshold <- 0.5

# Filter out outliers
cleaned_annual_isoforest <- cleaned_annual[outlier_scores < threshold, ]
write.csv(cleaned_annual_isoforest, "D:/Desktop/CSP-571-Proj/dataset/cleaned_annual_isoforest.csv", row.names = FALSE)
```

### Data Transformation

#### Normalization and Standardization

```{r}
# Load necessary libraries including dplyr
library(dplyr)

# Min-Max Normalization
cleaned_annual_isoforest <- cleaned_annual_isoforest %>%
  mutate(trip_duration_normalized = (trip_duration - min(trip_duration))/(max(trip_duration) - min(trip_duration)))

# Standardization
cleaned_annual_isoforest <- cleaned_annual_isoforest %>%
  mutate(trip_duration_standardized = (trip_duration - mean(trip_duration)) / sd(trip_duration))
```

#### Log Transformation

```{r}
# Log Transformation
cleaned_annual_isoforest <- cleaned_annual_isoforest %>%
  mutate(trip_duration_log = log(trip_duration + 1))
```

#### Date and Time Features

```{r}
# Load necessary libraries including dplyr and lubridate
library(dplyr)
library(lubridate)

# Update date and time features
cleaned_annual_isoforest <- cleaned_annual_isoforest %>%
  mutate(
    start_date = as.Date(started_at),
    start_month = lubridate::month(started_at, label = TRUE),
    start_day = day(started_at),
    start_hour = hour(started_at),
    start_minute = minute(started_at),
    end_date = as.Date(ended_at),
    end_month = lubridate::month(ended_at, label = TRUE),
    end_day = day(ended_at),
    end_hour = hour(ended_at),
    end_minute = minute(ended_at)
  )
```

#### Feature Engineering

```{r}
# Load necessary packages
library(dplyr)
library(tidyr)
library(lubridate)

# Assuming cleaned_annual_isoforest is your cleaned and preprocessed data

# Ensure rideable_type and member_casual are factors
cleaned_annual_isoforest <- cleaned_annual_isoforest %>%
  mutate(
    rideable_type = as.factor(rideable_type),
    member_casual = as.factor(member_casual)
  )


# Additional feature engineering
transformed_data <- transformed_data %>%
  mutate(
    trip_duration_hours = trip_duration / 60,
    time_of_day = case_when(
      start_hour >= 5 & start_hour < 12 ~ "Morning",
      start_hour >= 12 & start_hour < 17 ~ "Afternoon",
      start_hour >= 17 & start_hour < 21 ~ "Evening",
      TRUE ~ "Night"
    ),
    season = case_when(
      month(started_at) %in% c(12, 1, 2) ~ "Winter",
      month(started_at) %in% c(3, 4, 5) ~ "Spring",
      month(started_at) %in% c(6, 7, 8) ~ "Summer",
      month(started_at) %in% c(9, 10, 11) ~ "Fall"
    )
  )

# Display the transformed dataset
head(transformed_data)
write.csv(transformed_data, "D:/Desktop/CSP-571-Proj/dataset/transformed_data.csv", row.names = FALSE)

```
